Supplementary material for "Hack Weeks as a model for Data Science Education and Collaboration"

### 2018-01-07 sections/conclusions.tex

As introduced above, hack weeks serve multiple purposes, including dissemination of state-of-the-art technological advances through the scientific community, building collaborations between academic subdisciplines and fostering interdisciplinary research as well as  promoting open science and reproducibility.
Initial results from three events held in 2016 \textbf{and 2017} in three different fields (astronomy, geosciences and neurosciences) indicate that hack weeks succeed at all of these objectives, but that the measure of success is field-specific in that it depends to some degree on how much the concepts hack weeks promote were already adopted within the community.
Hack weeks are still a very young concept, and estimating the long-term impact of these events within the scientific communities they serve will require follow-up over multiple years to asses their effect on collaboration networks, career outcomes for early-career academics and adoption of new methods.
We have shown, however, that hack weeks provide an easy-to-implement, fairly low-cost method to introduce new technologies and methods into scientific fields on much shorter time scales than traditional teaching efforts can.

>>>>

As introduced above, hack weeks serve multiple purposes, including dissemination of state-of-the-art technological advances through the scientific community, building collaborations between academic subdisciplines and fostering interdisciplinary research as well as  promoting open science and reproducibility.
Initial results from three events held in 2016 and 2017 in three different fields (astronomy, geosciences and neurosciences) indicate that hack weeks succeed at all of these objectives, but that the measure of success is field-specific in that it depends to some degree on how much the concepts hack weeks promote were already adopted within the community.
Hack weeks are still a very young concept, and estimating the long-term impact of these events within the scientific communities they serve will require follow-up over multiple years to asses their effect on collaboration networks, career outcomes for early-career academics and adoption of new methods.
We have shown, however, that hack weeks provide an easy-to-implement, fairly low-cost method to introduce new technologies and methods into scientific fields on much shorter time scales than traditional teaching efforts can.
While we focus here on hack weeks in scientific fields, the concept could be extended to other areas, and is more generally useful in any area (1) where useful tools can be learned in short tutorials, (2) where results and outcomes can be produced on the timescale of a few days, and (3) that would benefit from collaborative approaches that cross traditional boundaries. Such areas could include for example the social sciences, music and art.

### 2018-01-07 sections/design.tex

Hack week outcomes depend strongly on the interest and engagement of participants.
Some attendees arrive with the goal of writing a scientific article.
Other attendees plan to learn a specific topic, such as machine learning, or to analyze a specific data using the tools covered in the tutorials.
This leads to a wide variety of project types from sandbox-style explorations to focused work efforts.
This breadth of possible outcomes makes it difficult to design for all possible participant goals, and calls for adaptive, flexible leadership among hack week organizers.
\textbf{The large variety in participant backgrounds and experiences--and the resulting range in personalities of attendees--requires careful, active facilitation of both taught and project components of a hack week. Community building is a core component of a hack week, and designs need to include both very strong personalities and very shy participants. In particular the presence of impostor syndrome experienced by many participants must be taken into account during workshop design (see also supplementary materials for concrete suggestions).}
%can lead to an increase in the prevalence of impostor syndrome experienced by many participants (see also supplementary materials), and it

>>>>

Hack week outcomes depend strongly on the interest and engagement of participants.
Some attendees arrive with the goal of writing a scientific article.
Other attendees plan to learn a specific topic, such as machine learning, or to analyze a specific data using the tools covered in the tutorials.
This leads to a wide variety of project types from sandbox-style explorations to focused work efforts.
This breadth of possible outcomes makes it difficult to design for all possible participant goals, and calls for adaptive, flexible leadership among hack week organizers.
The large variety in participant backgrounds and experiences--and the resulting range in personalities of attendees--requires careful, active facilitation of both taught and project components of a hack week. Community building is a core component of a hack week, and designs need to include both very strong personalities and very shy participants. In particular the presence of impostor syndrome experienced by many participants must be taken into account during workshop design (see also supplementary materials for concrete suggestions.
%can lead to an increase in the prevalence of impostor syndrome experienced by many participants (see also supplementary materials), and it

### 2018-01-07 sections/intro.tex

%\section*{}
\label{sec:introduction}
\dropcap{A}s data become cheaper to gather and store, researchers have become increasingly reliant on computational workflows requiring skills in statistical modeling, machine learning and scalable computation. In addition, the recent reproducibility crises in several scientific fields has motivated the acquisition of skills in open science and the design of reproducible workflows \cite[e.g.][]{pashler2012,frye2015,gezelter2015,baker2016}.
Formal university curricula have been relatively slow to offer courses in these important topics: the slack in this area has often been picked-up by extra-curricular, ad-hoc efforts such as workshops \cite{demasi2017}.
Well-known examples are the Software and Data Carpentry workshops providing training in research computing skills through a volunteer instructor program \cite{b:wilson-swc-lessons-2016,teal2015data}.
At the same time, there has been a rise in the number of statistical and computational courses designed for specific scientific disciplines.
Examples include the \textit{Summer School in Statistics for Astronomers}\footnote{\url{http://astrostatistics.psu.edu/su16/}}, the Google Earth Engine User Summits\footnote{\url{https://events.withgoogle.com/google-earth-engine-user-summit-2017/}}, as well as a variety of project-focused (rather than pedagogical) meetings, such as the dotAstronomy meetings\footnote{\url{http://dotastronomy.com}}.
Shorter, but similar-spirit meetings have been held in conjunction with conferences, such as the Hack Days at the annual American Astronomical Society meetings, the Brainhack hackathons associated with the Organization for Human Brain Mapping and the Society for Neuroscience\cite{Cameron_Craddock2016-wc}, and a hackathon at the American Geophysical Union meeting\footnote{\url{http://onlinelibrary.wiley.com/doi/10.1002/2014EO480004/pdf}}. 
In general, events to date have been designed withing either a pedagogical or project-based framework.
Pedagogically-focused events follow a classic academic model where novices learn new skills from experts, while project-focused workshops emphasize collaborative activities using existing skills.
A disadvantage of the pedagogical model is that it tends to focus on a one-way flow of information from instructor to student, and can discount the potential contributions by students.
A disadvantage of the project model is the common perception that the week is designed for technical experts, which may discourage others from attending.
In 2014, we initiated an alternative model of ``Hack Weeks'' that aim to combine the best of each of these models.
These week-long events combine pedagogy (often focused on statistical and computational techniques) with hacks and creative projects, with the goal of encouraging collaboration and learning among people at various stages of their career.

>>>>

%\section*{}
\label{sec:introduction}
As data become cheaper to gather and store, researchers have become increasingly reliant on computational workflows requiring skills in statistical modeling, machine learning and scalable computation. In addition, the recent reproducibility crises in several scientific fields has motivated the acquisition of skills in open science and the design of reproducible workflows \cite[e.g.][]{pashler2012,frye2015,gezelter2015,baker2016}.
Formal university curricula have been relatively slow to offer courses in these important topics: the slack in this area has often been picked-up by extra-curricular, ad-hoc efforts such as workshops \cite{demasi2017}.
Well-known examples are the Software and Data Carpentry workshops providing training in research computing skills through a volunteer instructor program \cite{b:wilson-swc-lessons-2016,teal2015data}.
At the same time, there has been a rise in the number of statistical and computational courses designed for specific scientific disciplines.
Examples include the \textit{Summer School in Statistics for Astronomers}\footnote{\url{http://astrostatistics.psu.edu/su16/}}, the Google Earth Engine User Summits\footnote{\url{https://events.withgoogle.com/google-earth-engine-user-summit-2017/}}, as well as a variety of project-focused (rather than pedagogical) meetings, such as the dotAstronomy meetings\footnote{\url{http://dotastronomy.com}}.
Shorter, but similar-spirit meetings have been held in conjunction with conferences, such as the Hack Days at the annual American Astronomical Society meetings, the Brainhack hackathons associated with the Organization for Human Brain Mapping and the Society for Neuroscience\cite{Cameron_Craddock2016-wc}, and a hackathon at the American Geophysical Union meeting\footnote{\url{http://onlinelibrary.wiley.com/doi/10.1002/2014EO480004/pdf}}.
In general, events to date have been designed withing either a pedagogical or project-based framework.
Pedagogically-focused events follow a classic academic model where novices learn new skills from experts, while project-focused workshops emphasize collaborative activities using existing skills.
A disadvantage of the pedagogical model is that it tends to focus on a one-way flow of information from instructor to student, and can discount the potential contributions by students.
A disadvantage of the project model is the common perception that the week is designed for technical experts, which may discourage others from attending.
In 2014, we initiated an alternative model of ``Hack Weeks'' that aim to combine the best of each of these models.
These week-long events combine pedagogy (often focused on statistical and computational techniques) with hacks and creative projects, with the goal of encouraging collaboration and learning among people at various stages of their career.

### 2018-01-07 sections/participants.tex

\section*{Audience and Participant Selection}

Hack weeks differ from traditional conferences or summer schools in that knowledge transfer occurs across many levels of seniority, disciplinary boundaries, and novelty of the topics discussed.
In addition, a substantial amount of hack week content is generated during the event itself, requiring active participation from participants.
Therefore in order to maximize learning outcomes and collaborative exchanges, it is crucial that the participant selection process be carried out with considerable care.

@@

No matter the selection procedures used, we encourage organizers to critically examine their cohort selection, experiment with new approaches, and routinely evaluate the different steps in their procedure%.  For example, merit scores can be interrogated for shared biases, and one can plot the demographics both before and after the initial selection step to diagnose whether certain groups are preferentially selected or excluded
(see also the Supplementary materials for more details on how the previous hack weeks approached this problem).

%One way to maximize transparency in the selection process is to minimize human decision making steps that introduce biases, and to . 
%We work to achieve this by first assessing the merit of each candidate with respect to the overall goals of the hack week.
%We try to minimize bias in this step by blinding ourselves to a candidate's other attributes, including name and other personal information, and assess their candidacy based soley on questions asked specifically for this purpose.
%When doing this procedure for a large enough sample, it is unlikely that the resulting pool of acceptable candidates is smaller than the number of available spaces at the workshop.

>>>>

\section*{Audience and Participant Selection}
Hack weeks differ from traditional conferences or summer schools in that knowledge transfer occurs across many levels of seniority, disciplinary boundaries, and novelty of the topics discussed.
In addition, a substantial amount of hack week content is generated during the event itself, requiring active participation from participants.
Therefore in order to maximize learning outcomes and collaborative exchanges, it is crucial that the participant selection process be carried out with considerable care.

@@

No matter the selection procedures used, we encourage organizers to critically examine their cohort selection, experiment with new approaches, and routinely evaluate the different steps in their procedure%.  For example, merit scores can be interrogated for shared biases, and one can plot the demographics both before and after the initial selection step to diagnose whether certain groups are preferentially selected or excluded
(see also the Supplementary materials for more details on how the previous hack weeks approached this problem).
%One way to maximize transparency in the selection process is to minimize human decision making steps that introduce biases, and to . 
%We work to achieve this by first assessing the merit of each candidate with respect to the overall goals of the hack week.
%We try to minimize bias in this step by blinding ourselves to a candidate's other attributes, including name and other personal information, and assess their candidacy based soley on questions asked specifically for this purpose.
%When doing this procedure for a large enough sample, it is unlikely that the resulting pool of acceptable candidates is smaller than the number of available spaces at the workshop.

### 2018-01-07 sections/results.tex

\caption{{\bf Post-workshop surveys from three hack weeks}: participants in the 2016 astro-, geo- and neuro- hack weeks responded to questions assessing their experiences. \textbf{For GHW and NHW, response rates were 100\%, with $N_{\mathrm{GHW}}= 83$ and $N_{\mathrm{NHW}} = 86$, respectively. At AHW, the combined response rate for both years was 76\%, or $N_{\mathrm{AHW}} = 72$ out of $94$ total attendees. We report here about results in three different domains: the development of technical skills (a -- e), collaboration and learning (f,g), and shifts in attitudes towards reproducibility and open science (h -- j)}}
\label{fig:survey}
%\end{center}
\end{figure*}

@@

Focusing on the outcomes of astro-, geo- and neuro- hack weeks (AHW, GHW, NHW, respectively) from 2016 and 2017, we find that \textbf{most} participants self-report successful learning outcomes (AHW: 76\%, GHW: 89\%, NHW: 79\% for responses ``somewhat agree'', ``agree'' and ``strongly agree''; Figure \ref{fig:survey}, (a)).
The overwhelming majority of respondents at the hack weeks ($>95\%$ for all three events) believed that they learned things that improve their day-to-day research, and that attendance has made them a better scientist (Figure \ref{fig:survey}, (b, c) ).
\textbf{More specifically, we compared learning outcomes in data visualization (the only topic explicitly shared between all three hack weeks; Figure \ref{fig:survey}, (d, e). We asked participants to subjectively rate their knowledge before the hack week, and find skill levels to be broadly distributed. This is expected, given the goal of diversity during the selection stage. We find that most participants have positive learning outcomes at all three hack weeks for data visualization, but that outcomes vary strongly, as is expected, too, with a group that is this diverse. We also note that there might be systematic effects and biases due to the subjectivity of rating knowledge and learning outcomes.}

The majority of participants felt that they built valuable connections to other researchers (Figure \ref{fig:survey}, (g)), especially at Neuro Hack Week, where more than 64\% of participants strongly agreed with this statement. 
Because peer learning is a major mode of knowledge transfer at hack weeks, we asked participants whether they taught other participants.
We find that again a majority agrees with this statement to some degree (AHW: 79\%, GHW: 69\%, NHW: 75\%; Figure \ref{fig:survey}, (f)), though responses are not as unequivocal as they are in some of the other categories.
Similarly to learning outcomes, this is expected given the range of skill levels at the workshops. \textbf{It is notable, however, that we find no correlation between the response to this question and career stage for any of the events surveyed here. Participants at all career stages report similar engagement in teaching. This suggests there is some evidence for our hypothesis that compared to e.g.\ a traditional summer school, a hack week is less hierarchical and encourages lateral knowledge transfer between participants at different career stages.  We similarly find no correlations with gender identity or race/ethnicity.}

We find that the hack weeks have been largely successful at efforts to promote positive attitudes towards reproducibility and open science: at all three events, more than 85\% of all participants (AHW: 86\%, GHW: 94\%, NHW: 95\%; Figure \ref{fig:survey}, (h)) put code or data created at the hack week into a public repository, while a substantially smaller fraction of participants followed \textbf{a regular practice of publishing code} before the event (Figure \ref{fig:survey}, (i)).
The overall behaviour reflects how general conventions and attitudes differ in different fields.
Astronomy shows the largest degree of openness toward open science, whereas our results indicate that open science is still fairly uncommon in the geosciences, with neuroscience falling in between.
%This implies that hack weeks can have the highest impacts in field where a priori engagement in reproducibility efforts is low and significant progress can be made towards changing researchers' attitudes during a collaborative workshop.
Similar attitudes are reflected when asking whether the hack week has made participants more comfortable with open science (Figure \ref{fig:survey}, (j)): again, geoscience shows the large improvement with over 97\% agreeing with this statement to some degree, followed by neuroscience (95\%) and astronomy (72\%).
Overall, our results indicate that hack weeks are effective at addressing persisting doubts about making research open and reproducible. \textbf{While the focus on open science is not necessarily a required component of a hack week, it aligns naturally with many of the goals and values commonly promoted at hack weeks like open-source software and data sharing. In some fields, especially where ethical issues around data sharing and privacy are relevant, this might not be a desired focus of the hack week, or might be replaced or augmented by a discussion of ethical considerations.}

>>>>

\caption{{\bf Post-workshop surveys from three hack weeks}: participants in the 2016 astro-, geo- and neuro- hack weeks responded to questions assessing their experiences. For GHW and NHW, response rates were 100\%, with $N_{\mathrm{GHW}}= 83$ and $N_{\mathrm{NHW}} = 86$, respectively. At AHW, the combined response rate for both years was 76\%, or $N_{\mathrm{AHW}} = 72$ out of $94$ total attendees. We report here about results in three different domains: the development of technical skills (a -- e), collaboration and learning (f,g), and shifts in attitudes towards reproducibility and open science (h -- j)}
\label{fig:survey}
%\end{center}
\end{figure*}

@@

Focusing on the outcomes of astro-, geo- and neuro- hack weeks (AHW, GHW, NHW, respectively) from 2016 and 2017, we find that most participants self-report successful learning outcomes (AHW: 76\%, GHW: 89\%, NHW: 79\% for responses ``somewhat agree'', ``agree'' and ``strongly agree''; Figure \ref{fig:survey}, (a)).
The overwhelming majority of respondents at the hack weeks ($>95\%$ for all three events) believed that they learned things that improve their day-to-day research, and that attendance has made them a better scientist (Figure \ref{fig:survey}, (b, c) ).
More specifically, we compared learning outcomes in data visualization (the only topic explicitly shared between all three hack weeks; Figure \ref{fig:survey}, (d, e). We asked participants to subjectively rate their knowledge before the hack week, and find skill levels to be broadly distributed. This is expected, given the goal of diversity during the selection stage. We find that most participants have positive learning outcomes at all three hack weeks for data visualization, but that outcomes vary strongly, as is expected, too, with a group that is this diverse. We also note that there might be systematic effects and biases due to the subjectivity of rating knowledge and learning outcomes.

The majority of participants felt that they built valuable connections to other researchers (Figure \ref{fig:survey}, (g)), especially at Neuro Hack Week, where more than 64\% of participants strongly agreed with this statement.
Because peer learning is a major mode of knowledge transfer at hack weeks, we asked participants whether they taught other participants.
We find that again a majority agrees with this statement to some degree (AHW: 79\%, GHW: 69\%, NHW: 75\%; Figure \ref{fig:survey}, (f)), though responses are not as unequivocal as they are in some of the other categories.
Similarly to learning outcomes, this is expected given the range of skill levels at the workshops. It is notable, however, that we find no correlation between the response to this question and career stage for any of the events surveyed here. Participants at all career stages report similar engagement in teaching. This suggests there is some evidence for our hypothesis that compared to e.g.\ a traditional summer school, a hack week is less hierarchical and encourages lateral knowledge transfer between participants at different career stages.  We similarly find no correlations with gender identity or race/ethnicity.

We find that the hack weeks have been largely successful at efforts to promote positive attitudes towards reproducibility and open science: at all three events, more than 85\% of all participants (AHW: 86\%, GHW: 94\%, NHW: 95\%; Figure \ref{fig:survey}, (h)) put code or data created at the hack week into a public repository, while a substantially smaller fraction of participants followed a regular practice of publishing code before the event (Figure \ref{fig:survey}, (i)).
The overall behaviour reflects how general conventions and attitudes differ in different fields.
Astronomy shows the largest degree of openness toward open science, whereas our results indicate that open science is still fairly uncommon in the geosciences, with neuroscience falling in between.
%This implies that hack weeks can have the highest impacts in field where a priori engagement in reproducibility efforts is low and significant progress can be made towards changing researchers' attitudes during a collaborative workshop.
Similar attitudes are reflected when asking whether the hack week has made participants more comfortable with open science (Figure \ref{fig:survey}, (j)): again, geoscience shows the large improvement with over 97\% agreeing with this statement to some degree, followed by neuroscience (95\%) and astronomy (72\%).
Overall, our results indicate that hack weeks are effective at addressing persisting doubts about making research open and reproducible. While the focus on open science is not necessarily a required component of a hack week, it aligns naturally with many of the goals and values commonly promoted at hack weeks like open-source software and data sharing. In some fields, especially where ethical issues around data sharing and privacy are relevant, this might not be a desired focus of the hack week, or might be replaced or augmented by a discussion of ethical considerations.
